---
title: "Linear Regression - Predicting Individual Medical Costs Billed by Health Insurance"
author: "Name: Erick Rivadeneira"
subtitle: "Date: 04/29/2024"
output: html_document
---

## Introduction
"The United States’ national health expenditure (NHE) grew 5.8% to $3.2 trillion in 2015 (i.e., $9,990 per person), which accounted for 17.8% of the nation’s gross domestic product (GDP). In seeking to control these unsustainable increases in healthcare costs, it is imperative that healthcare organizations can predict the likely future costs of individuals, so that care management resources can be efficiently targeted to those individuals at highest risk of incurring significant costs. Key stakeholders in these efforts to manage healthcare costs include health insurers, employers, society, and increasingly healthcare delivery organizations due to the transition from fee-for-service payment models to value-based payment models. For any given individual, insurers generally have the most comprehensive information on healthcare costs as they pay for care delivered across various healthcare delivery organizations.

Predicting healthcare costs for individuals using accurate prediction models is important for various stakeholders beyond health insurers, and for various purposes. For health insurers and increasingly healthcare delivery systems, accurate forecasts of likely costs can help with general business planning in addition to prioritizing the allocation of scarce care management resources. Moreover, for patients, knowing in advance their likely expenditures for the next year could potentially allow them to choose insurance plans with appropriate deductibles and premiums."

Excerpt from https://www-ncbi-nlm-nih-gov.proxy-um.researchport.umd.edu/pmc/articles/PMC5977561/

## Exploratory Data Analysis
From the patient data in which this study has collected, 1338 insured patients were analyzed into 7 variables: age of primary beneficiary, sex, bmi, number of children covered by their plan, smoking status, region of the US in which they live and individual medical costs billed by health insurance (in dollars). A majority of our patients share the following demographics: On average primary beneficiaries are aged 39 years old with a standard deviation of 14 years, with the majority of patients being men (51%) with an average bmi of 30.7±6.1, and having 0 children covered by their plan (43%), 79% of them self identifying themselves as non-smokers, with a majority of our patients living in the southeast region of the US (27%), and resulting in an average insurance charge of $13,270.42 ± 12110.01.
```{r warning=FALSE, message=FALSE, results="hide", include=FALSE}
library(tidyverse)
library(RNHANES)
library(ggplot2)
#import file
insurance <-read.csv("insurance.csv", header=TRUE)
#find number of observations and variables
dim(insurance)
#percentage of observations in each category
#age- continuous
mean(insurance$age, na.rm=T)
sd(insurance$age, na.rm=T)
#sex- categorical
prop.table(table(insurance$sex))*100
#bmi- continuous
mean(insurance$bmi, na.rm=T)
sd(insurance$bmi, na.rm=T)
#children- categorical
prop.table(table(insurance$children))*100
#smoker- categorical
prop.table(table(insurance$smoker))*100
#region- categorical
prop.table(table(insurance$region))*100
#charges- continuous
mean(insurance$charges, na.rm=T)
sd(insurance$charges, na.rm=T)

```

## Data Visualization
With regards to this data, 6 visualizations were made.**Figure 1** is a boxplot that displays the distribution of medical costs billed by health insurance by sex. The median medical costs between male and females are the same (about 9500) as well as the Q1 value (about 5000) as well as the minimum value (about 2000). However, they differ from the Q3 medical costs (15000 female and 19000 male), as well as the maximum cost (66000 female and 65000 male).**Figure 2** is a boxplot that displays the distribution of medical costs billed by health insurance by number of children covered. No children can be covered and the maximum that can be covered is up to 5 according to this data, with each category of number of children from 0 to 5, the data differs with regards to minimum, maximum, Q1, median, and Q3. For insured patients that have 0 children, the minimum charge is 2000, the Q1 is 3000, the median is 10000, Q3 is 15000, and the maximum is 65000. For insured people that have 1 child covered, the minimum charge is 3000, the Q1 is 5000, the median is 8000, Q3 is 15000, and the maximum is 59000. For insured patients that have 2 children covered, the minimum charge is 4000, the Q1 is 7000, the median is 9000, the Q3 is 20000 and the maximum is 50000. For insured patients that have 3 children covered, the minimum is 4500, the Q1 is 7500, the median is 10500, the Q3 is 19000, and the maximum is 60000. For insured patients that have 4 children covered, the minimum is 6000, the Q1 is 8000, the median is 11000, the Q3 is 16000 and the maximum is 40500. For insured patients that have 5 children covered, the minimum is 6500, the Q1 is 6500, the median is 9000, the Q3 is 10000 and the maximum is 19000. **Figure 3** is a boxplot that displays the distribution of medical costs billed by health insurance by smoker status. For Patients that do smoke the minimum charge is 3000, the Q1 is 5000, the median is 80000, the Q3 is 11000, and the maximum is 37000. **Figure 4** is a boxplot that displays the distribution of medical costs billed by health insurance by the geographic region of the US which the insured patient lives. For patients that live in the northeast, the minimum charge is 3000, the Q1 is 5000, the median is 10000, the Q3 is 17000, the maximum is 59000.For patients that live in the northwest, the minimum charge is 3000 which is the same as those that live in the northeast, the Q1 is 4500, the median is 9000, the Q3 is 15000, and the maximum is 60000.For patients that live in the southeast, the minimum charge is 2000, the Q1 is 4000, the median is 9500, the Q3 is 19500, and the maximum is 65000. For patients that live in the southwest, the minimum charge is 2000 which is the same as those that live in the southeast, the Q1 is 4500 which is the same as the northwest, the median is 9000 which is the same as the northwest, the Q3 is 14000, and the maximum is 52000. **Figure 5** is a scatter plot that shows a positive linear relationship between BMI and the Health Insurance Charges. It is notable to include that the relationship has a weak positive correlation due to the correlation coefficient being 0.2. **Figure 6** is a scatter plot that shows a  positive linear relationship between Age and Health Insurance Charges. It is notable to include that the relationship has a weak positive correlation due to the correlation coefficient being 0.3.

```{r warning=FALSE, message=FALSE, echo=FALSE}

#Sex
ggplot(data=insurance) + aes(x=sex, y=charges, fill=sex) +  geom_boxplot()+
   labs(title="Figure 1: BoxPlot of Individual medical costs billed by health insurance, by Sex",x="Sex(Gender)", y="Health Insurance Charges (in Dollars)",fill="Sex") + theme(plot.title = element_text(hjust = 0.5))
#Children
ggplot(data=insurance) + aes(x=factor(children), y=charges, fill=children) +  geom_boxplot()+
   labs(title="Figure 2: BoxPlot of Individual medical costs billed by health insurance, by number of Children covered",x="Number of Children (#)", y="Health Insurance Charges (in Dollars)",fill="Number of Children (#)") + theme(plot.title = element_text(hjust = 0.5))
#Smoker
ggplot(data=insurance) + aes(x=smoker, y=charges, fill=smoker) +  geom_boxplot()+
   labs(title="Figure 3: Individual medical costs billed by health insurance, by Smoking Status",x="Smoker (Yes or No)", y="Health Insurance Charges (in Dollars)",fill="Smoker Status") + theme(plot.title = element_text(hjust = 0.5))
#Region
ggplot(data=insurance) + aes(x=region, y=charges, fill=region) +  geom_boxplot()+
   labs(title="Figure 4: Individual medical costs billed by health insurance by Geographic Region",x="Geographic Region (in the US)", y="Health Insurance Charges (in Dollars)",fill="Region(US)") + theme(plot.title = element_text(hjust = 0.5))


#BMI Scatter Plot
ggplot(data=insurance) + aes(x=bmi, y=charges) +  geom_point() + geom_smooth(method=lm) +
   labs(title="Individual medical costs billed by health insurance vs BMI",x="Body Mass Index (BMI)", y="Health Insurance Charges (in Dollars)") + theme(plot.title = element_text(hjust = 0.5))
# correlation coefficient
cor.test(insurance$bmi, insurance$charges, method = "pearson", conf.level = 0.95)
 
 
#Age
ggplot(data=insurance) + aes(x=age, y=charges) +  geom_point() + geom_smooth(method=lm) +
   labs(title="Individual medical costs billed by health insurance vs Age",x="Age (In Years)", y="Health Insurance Charges (in Dollars)") + theme(plot.title = element_text(hjust = 0.5))
# correlation coefficient
cor.test(insurance$age, insurance$charges, method = "pearson", conf.level = 0.95)

```

## Multiple Linear Regression Results
Health insurance charges are -128.16 dollars lower in females than males, holding all other variables in the model constant. Health insurance charges are 1116.04 dollars higher in insured patient with 5 children than insured patients with no children. Health insurance charges are 2947.37 dollars higher in insured patients with 4 children than insured patients with no children. Health insurance charges are 964.34 dollars higher in insured patients with 3 children than insured patients with no children. Health insurance charges are 1635.78 dollars higher in insured patients with 2 children than insured patients with no child. Health insurance charges are 390.98 higher in insured patients with 1 child than insured patients with no child. A one-unit change in age corresponds to an 257.19 average increase in health insurance charges, all other variables held constant. Health insurance charges are 1116.04 dollars higher in insured patient with 5 children than insured patients with 4 children.Health insurance charges are 23836.41 dollars higher in non-smokers than smokers, holding all other variables in the model constant. Health insurance charges are 1116.04 dollars higher in insured patient with 5 children than insured patients with 4 children. Health insurance charges are 380.04 dollars lower in other regions than people who live in the northwest, holding all other variables in the model constant. Health insurance charges are 1033.14 dollars lower in other regions than people who live in the southeast, holding all other variables in the model constant. Health insurance charges are 952.89 dollars lower in other regions than people who live in the southwest, holding all other variables in the model constant. A one-unit change in BMI corresponds to an 336.91 average increase in health insurance charges, all other variables held constant.

```{r warning=FALSE, message=FALSE, results="hide", include=FALSE}

mlr <- lm(charges ~ sex + factor(children) + age + smoker + region + bmi, data = insurance)
summary(mlr)

```

## Linear Regression Assumptions
The linearity of data is assessed by the **Residuals vs Fitted**graph and our data appears to violate this assumption because there are no distinct patterns that indicate a linear relationship. Normality of residuals is assessed by the **Normal Q-Q** plot and our data appears to violate this assumption because the dots don't exclusively follow the dashed line completely. Homogeneity of residuals variance is assessed by the **Scale-Location (or Spread-Location)** plot and it appears our data appears to violate this assumption because our red line has a positive slope to it when it should be flat. Violating all these regression assumptions may mean our results are biased or inefficient.
```{r warning=FALSE, message=FALSE, echo=FALSE}
plot(mlr)
```

